Constructing a universe for the setoid model

### 2021-01-21 paper.tex
Fix gross LaTeX misuse: * \parallel is a relation symbol, not to be used for the truncation operator * replaced '...' in formulas by \ldots
Use macro \def with vertically centered colon for defining equalities.

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
%\usepackage[bw]{agda}

@@

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
\vspace{-0.2em}
\begin{align*}
  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

@@

Notably, equality types in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%% %
%% \begin{align*}
%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
%%   & \Prop^\sim\ p\ P\ Q = ...
%% \end{align*}

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t :\equiv
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$. \\
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.

@@

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
\end{alignat*}

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\inUU$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:
\[
(A, a) \sim_{\setoidU} (B, b) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
\]

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash x \sim_{\El} y :\equiv \ ?
\end{align*}

In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
\pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
$\blank\vdash\blank\sim_{\El}\blank$ in line with how we defined $\setoidU$ and $\El$.

@@

We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
\Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
:\equiv \Sigma\ (A_0 :
\textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straightforward:
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
\end{alignat*}

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
\usepackage{mathtools}
%\usepackage[bw]{agda}

@@

% for definitions
\newcommand{\defeq}{\mathrel{\vcentcolon\equiv}}

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
\vspace{-0.2em}
\begin{align*}
  \mathbf{1}_\mProp & \defeq \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & \defeq \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & \defeq \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

@@

Notably, equality types in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%% %
%% \begin{align*}
%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)\ldots \\
%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = \ldots \\
%%   & \Prop^\sim\ p\ P\ Q = \ldots
%% \end{align*}

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t \defeq
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$. \\
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

$\U$ and $\El$ can be given by $\U \defeq \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El \defeq \pi_1$.

@@

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty \defeq \reflu \\
  & \ldots \\
  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
  & \ldots \\
  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
  & \setoidU && \defeq \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El \defeq \pi_1
\end{alignat*}

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\inUU$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:
\[
(A, a) \sim_{\setoidU} (B, b) \defeq \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
\]

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash x \sim_{\El} y \defeq \ ?
\end{align*}

In the definition above, $p$ has type $\| \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (\ldots) \|$. If the type was not propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y \defeq
\pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
$\blank\vdash\blank\sim_{\El}\blank$ in line with how we defined $\setoidU$ and $\El$.

@@

We can recover the original inductive-inductive type as $\textsf{Con} \defeq
\Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
\defeq \Sigma\ (A_0 :
\textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straightforward:
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && \defeq (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) && \defeq ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) && \defeq (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && \defeq
    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
\end{alignat*}
